### 2015-04-06 cug_paper/02_design.tex
Replace fig 1 with two figures

\subsection{Overview}
\label{subsec:overview}
This idea was tested using a synthetic benchmark application written by the authors.  The benchmark consists of two executables: the main application which uses MPI and simulates the I/O patterns of a typical HPC application and a daemon process that is started on each compute node.  The daemon has two tasks: manage the GPU memory and copy data from GPU memory out to a file.  Note that the daemon does \emph{not} copy the data into the GPU memory; that is handled by the individual ranks of the main application.   There needs to be exactly one daemon process on each compute node regardless of how many MPI ranks are running on each node.  \footnote{Due to the limitations of aprun, starting this daemon is somewhat involved.  See section \ref{subsec:cray-specific}.}

With the daemon process running on each node, the application begins its main execution.  In this case, since it's just a synthetic 
I/O benchmark, it performs no actual computation; it simply writes a specified amount of data and then sleeps for a specified amount of time.  For these tests, the sleep time was deliberately calculated to allow enough time for the data in GPU memory to drain.
\subsection{Communication Between the Application and Daemon}
\label{subsec:comm}

Communication between the main application and the daemon happens over two different channels.  Short messages for coordination between the application and the daemon are handled using the Common Communication Interface (CCI).\cite{atchley11:cci}  Bulk data is copied to and from GPU memory using \texttt{cudaMemcpy()}.

\begin{figure*}[hp]
\centerline{
  \subfloat[Application sends write request message to daemon]{
    \includegraphics[width=2.5in]{figures/Data_Movement/Slide09}%
    \label{fig:comm_a}}
  \hfil
  \subfloat[Daemon allocates GPU memory and converts pointer into IPC handle]{
    \includegraphics[width=2.5in]{figures/Data_Movement/Slide10}%
    \label{fig:comm_b}}}
\centerline{
  \subfloat[Daemon sends reply message with IPC handle]{
    \includegraphics[width=2.5in]{figures/Data_Movement/Slide11}%
    \label{fig:comm_c}}
  \hfil
  \subfloat[Application writes data to GPU memory]{
    \includegraphics[width=2.5in]{figures/Data_Movement/Slide12}%
    \label{fig:comm_d}}}    
\centerline{
  \subfloat[Application sends write complete message to daemon]{
    \includegraphics[width=2.5in]{figures/Data_Movement/Slide13}%
    \label{fig:comm_e}}
  \hfil
  \subfloat[Daemon copies data from GPU memory to system memory]{
    \includegraphics[width=2.5in]{figures/Data_Movement/Slide14}%
    \label{fig:comm_f}}} 
\centerline{
  \subfloat[Daemon writes data to filesystem]{
    \includegraphics[width=2.5in]{figures/Data_Movement/Slide15}%
    \label{fig:comm_g}}} 
\caption{Communications steps}
\label{fig:communications}
\end{figure*}

Writing data from the application is a seven step process.

When it's time to write, the application sends a short message to the daemon using CCI.  This message is a request to write data, and contains the amount of data and the starting offset into the output file.\footnote{For this benchmark, the daemon names output files based on the MPI rank of the requestor.  For production code, something more flexible will obviously be needed.}(Figure \ref{fig:comm_a})

When the daemon receives this message, it will attempt to allocate memory on the GPU.  Assuming the allocation succeeds, the daemon will use \texttt{cudaIpcGetMemhandle()} to create a memory handle for the allocation. (Figure \ref{fig:comm_b})

The daemon will then reply to the application with the size of memory that was actually allocated and the memory handle that the application can use to access the GPU memory.  Note that the size value may be less than what the application asked for. (Figure \ref{fig:comm_c})

Upon receiving the reply, the application opens the memory handle and converts it back to a standard pointer.  The application then uses \texttt{cudaMemcpy()} to copy the specified amount of data up to the GPU memory. (Figure \ref{fig:comm_d})

When the copy has completed, the application closes the memory handle and sends a message back to the daemon saying that this particular write has completed.  (Figure \ref{fig:comm_e})  If the application has more data to write, it can repeat the previous steps, or else it can continue with its calculations.

Note that at this point the data has not made it out to the filesystem yet; it's just sitting in GPU memory.

@@

The daemon maintains one or more background threads that are responsible for copying data out of GPU memory into normal system memory and then writing that data to the filesystem.  These threads allow the writes to the filesystem to happen asynchronously while the main thread handles all the CCI messages.  Since it's impossible to write directly from GPU memory to the filesystem, each thread will copy a block of data from GPU memory into a buffer in main memory.  (Figure \ref{fig:comm_f})  It will then write the data to the output file at the location specified in the original request message.  (Figure \ref{fig:comm_g})  Once data has been written to the filesystem, the write thread free's the GPU memory so that it can be used in another request.

@@

There are a few peculiarities of the Cray environment that need to be taken in to account in order for this software to work properly.  The first peculiarity has to do with limiting access to the GPU and associated hardware.  The CUDA runtime can be configured to limit access to the GPU to a single thread, to multiple threads from a single process or to multiple threads from multiple processes.  On Titan, if no special options are given to qsub, the CUDA runtime will be configured for a single thread.  Since software described in this paper requires access to the GPU from multiple processes, users must switch to the appropriate compute mode.  This is done by passing the flag "-l feature=gpudefault" to qsub.\footnote{It may seem strange to have to explicitly switch to a mode called 'gpudefault'.  The naming scheme comes from the CUDA documentation.  On a standard CUDA install, the GPU would in fact default to allowing access from multiple processes.  Titan is configured differently, however.  Thus the need to explicitly tell qsub to switch the GPU's to the `default' mode.}

The second peculiarity has to do with actually starting the daemon process.  Normally, executables are started on the compute nodes by aprun.  However, aprun is designed to start a single executable, not to start two different executables.  Even in MPMD mode, it won't start two different executables on the same node.  The solution is to have one rank on each node call \texttt{fork()} and then \texttt{execve()}.  Since MPI purposely hides the compute topology from the application, deciding exactly which ranks call fork() takes a little more work.  Specifically, all ranks will attempt to open a file using the \texttt{O\_CREAT} and 
\texttt{O\_EXCL} flags.  The name of the file is taken from the compute node's host name.  This combination of flags and filename ensures that exactly one rank on each node successfully opens the file.  The ranks that succeed will then start the daemon processes.

@@

It's debatable whether this sort of load balancing is necessary or even desirable.  Indeed, preliminary testing seemed to indicate that it might be faster to let the daemon allocate  as much GPU memory as the client requested (assuming there was enough free).  However, that same testing showed that one of the CUDA functions - \texttt{cudaIpcOpenMemHandle()} - became unreliable once GPU memory allocations exceeded 512MB.  Until the nature of that problem could be determined, it was decided to limit allocations to 16MB.

>>>>

\subsection{Overview}
\label{subsec:overview}
To test this idea, we created a synthetic benchmark application that consists of two executables: the main application which uses MPI and simulates the I/O patterns of a typical HPC application and a daemon process that is started on each compute node.  The daemon has two tasks: manage the GPU memory and copy data from GPU memory out to a file.  Note that the daemon does \emph{not} copy the data into the GPU memory; that is handled by the individual ranks of the main application. There needs to be exactly one daemon process on each compute node regardless of how many MPI ranks are running on each node.\footnote{Due to the limitations of aprun, starting this daemon is somewhat involved.  See section \ref{subsec:cray-specific}.}

With the daemon process running on each node, the application begins its main execution.  In this case, since it is just a synthetic 
I/O benchmark, it performs no actual computation; it simply writes a specified amount of data and then sleeps for a specified amount of time.  For these tests, the sleep time was deliberately calculated to allow enough time for the data in GPU memory to drain.
\subsection{Communication Between the Application and Daemon}
\label{subsec:comm}

The main application and daemon coordinate with each other using short messages provided by the Common Communication Interface (CCI).\cite{atchley11:cci} Bulk data is copied to and from GPU memory using \texttt{cudaMemcpy()}.

\begin{figure}
\includegraphics[trim = .75in .75in .75in .75in, clip, width=0.5\textwidth]{figures/Data_Movement/Slide20}%
\caption{Buffering data to GPU memory}
\label{fig:buffer_write}
\end{figure}

\begin{figure}
\includegraphics[trim = .75in .75in .75in .75in, clip, width=0.5\textwidth]{figures/Data_Movement/Slide21}%
\caption{Draining data from GPU memory}
\label{fig:buffer_drain}
\end{figure}

%\begin{figure*}[hp]
%\centerline{
%  \subfloat[Application sends write request message to daemon]{
%    \includegraphics[width=2.5in]{figures/Data_Movement/Slide09}%
%    \label{fig:comm_a}}
%  \hfil
%  \subfloat[Daemon allocates GPU memory and converts pointer into IPC handle]{
%    \includegraphics[width=2.5in]{figures/Data_Movement/Slide10}%
%    \label{fig:comm_b}}}
%\centerline{
%  \subfloat[Daemon sends reply message with IPC handle]{
%    \includegraphics[width=2.5in]{figures/Data_Movement/Slide11}%
%    \label{fig:comm_c}}
%  \hfil
%  \subfloat[Application writes data to GPU memory]{
%    \includegraphics[width=2.5in]{figures/Data_Movement/Slide12}%
%    \label{fig:comm_d}}}    
%\centerline{
%  \subfloat[Application sends write complete message to daemon]{
%    \includegraphics[width=2.5in]{figures/Data_Movement/Slide13}%
%    \label{fig:comm_e}}
%  \hfil
%  \subfloat[Daemon copies data from GPU memory to system memory]{
%    \includegraphics[width=2.5in]{figures/Data_Movement/Slide14}%
%    \label{fig:comm_f}}} 
%\centerline{
%  \subfloat[Daemon writes data to filesystem]{
%    \includegraphics[width=2.5in]{figures/Data_Movement/Slide15}%
%    \label{fig:comm_g}}} 
%\caption{Communications steps}
%\label{fig:communications}
%\end{figure*}

Writing data from the application is a seven step process. When it is time to write, the application sends a short message to the daemon using CCI.  This message is a request to write data, and contains the amount of data and the starting offset into the output file.\footnote{For this benchmark, the daemon names output files based on the MPI rank of the requestor.  For production code, something more flexible will obviously be needed.} (Figure \ref{fig:buffer_write}a)

When the daemon receives this message, it will attempt to allocate memory on the GPU.  Assuming the allocation succeeds, the daemon will use \texttt{cudaIpcGetMemhandle()} to create a memory handle for the allocation. (Figure \ref{fig:buffer_write}b)

The daemon will then reply to the application with the size of memory that was actually allocated and the memory handle that the application can use to access the GPU memory.  Note that the size value may be less than what the application asked for. (Figure \ref{fig:buffer_write}c)

Upon receiving the reply, the application opens the memory handle and converts it back to a standard pointer.  The application then uses \texttt{cudaMemcpy()} to copy the specified amount of data up to the GPU memory. (Figure \ref{fig:buffer_write}d)

When the copy has completed, the application closes the memory handle and sends a message back to the daemon saying that this particular write has completed.  (Figure \ref{fig:buffer_write}e)  If the application has more data to write, it can repeat the previous steps, or else it can continue with its calculations.

Note that at this point the data has not made it out to the filesystem yet; it is just sitting in GPU memory.

@@

The daemon maintains one or more background threads that are responsible for copying data out of GPU memory into normal system memory and then writing that data to the filesystem.  These threads allow the writes to the filesystem to happen asynchronously while the main thread handles all the CCI messages.  Since it is impossible to write directly from GPU memory to the filesystem, each thread will copy a block of data from GPU memory into a buffer in main memory.  (Figure \ref{fig:buffer_drain}f)  It will then write the data to the output file at the location specified in the original request message.  (Figure \ref{fig:buffer_drain}g)  Once data has been written to the filesystem, the write thread frees the GPU memory so that it can be used in another request.

@@

There are a few peculiarities of the Cray environment that need to be taken in to account in order for this software to work properly.  The first peculiarity has to do with limiting access to the GPU and associated hardware.  The CUDA runtime can be configured to limit access to the GPU to a single thread, to multiple threads from a single process or to multiple threads from multiple processes.  On Titan, if no special options are given to qsub, the CUDA runtime will be configured for a single thread.  Since software described in this paper requires access to the GPU from multiple processes, users must switch to the appropriate compute mode.  This is done by passing the flag "-l feature=gpudefault" to qsub.\footnote{It may seem strange to have to explicitly switch to a mode called `gpudefault'.  The naming scheme comes from the CUDA documentation.  On a standard CUDA install, the GPU would in fact default to allowing access from multiple processes.  Titan is configured differently, however.  Thus the need to explicitly tell qsub to switch the GPU's to the `default' mode.}

The second peculiarity has to do with actually starting the daemon process.  Normally, executables are started on the compute nodes by aprun.  However, aprun is designed to start a single executable, not to start two different executables.  Even in MPMD mode, it will not start two different executables on the same node.  The solution is to have one rank on each node call \texttt{fork()} and then \texttt{execve()}.  Since MPI purposely hides the compute topology from the application, deciding exactly which ranks call fork() takes a little more work.  Specifically, all ranks will attempt to open a file using the \texttt{O\_CREAT} and 
\texttt{O\_EXCL} flags.  The name of the file is taken from the compute node's host name.  This combination of flags and filename ensures that exactly one rank on each node successfully opens the file.  The ranks that succeed will then start the daemon processes.

@@

It is debatable whether this sort of load balancing is necessary or even desirable.  Indeed, preliminary testing seemed to indicate that it might be faster to let the daemon allocate  as much GPU memory as the client requested (assuming there was enough free).  However, that same testing showed that one of the CUDA functions - \texttt{cudaIpcOpenMemHandle()} - became unreliable once GPU memory allocations exceeded 512MB.  Until the nature of that problem could be determined, it was decided to limit allocations to 16MB.
